BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2014-07-09 Paper/BOAST.tex
updated paper

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

@@

Developed by physicist, chemist, meteorologist. Usually in FORTRAN for
historical and performance reason. Codes can be quite huge (several
thousands lines of code) with lots of functionalities. Nonetheless they are
usually based on computing kernels. Computing kernels are resource
intensive well defined parts of a program with a limited set of input and
output. Those kernels are the time consuming part of an HPC application.

@@

\section{Use Cases}

  \subsection{BigDFT}

  \subsection{SPECFEM3D}

\section{Using Code Generation in Application Autotuning}

  \subsection{General Principles}

  \subsection{Our Contribution: BOAST}
  \cite{videau2013boast}
     + dumper pour tests de non régression

\section{Creating an Auto-Tuned Convolution Library for BigDFT using BOAST}

\section{Porting SPECFEM3D to OpenCL using BOAST}

>>>>

\usepackage{lineno,hyperref}
\usepackage{listings}
\usepackage{color}
\usepackage{url}
\modulolinenumbers[5]

@@

Developed by physicist, chemist, meteorologist. Usually in FORTRAN for
historical and performance reason. Codes can be quite huge (several
thousands lines of code) with lots of functionalities. Nonetheless they are
usually based on computing kernels. Computing kernels are resource
intensive and well defined parts of a program with a limited set of input
and output. Those kernels are the time consuming part of an HPC
application. They are thus the prime target for optimization.

In Section~\ref{use_cases} we will present two HPC application that we used
as use cases: SPECFEM3D and BigDFT. They are both based on computing
kernels.

@@

The problematic here is to obtain computing kernels that present good
performances on the architectures encountered by the application while
still being portable after the optimization process took place. Indeed the
application might encounter one of the other architectures presented
before. Investing manpower to optimize the application for a new
architecture is reasonable, suffering hindrance from previous optimization
work is not. Thus optimizations have to be as orthogonal as possible from
one another so as to be easily activated and deactivated.

If this paradigm is followed by developers then they will rapidly be
confronted with a huge optimization space to search. They will need to
be able to test easily the performance impact of the chosen optimizations
without running the full application. The same reasoning implies that
kernels should be tested for non regression without running the full
application.

\section{BOAST: Using Code Generation in Application Autotuning}

  \cite{videau2013boast}

  \subsection{Kernel Description Language}

  \subsection{BOAST Runtime}
     + dumper pour tests de non régression

\section{Use Cases}
\label{use_cases}

  \subsection{Creating an Auto-Tuned Convolution Library for BigDFT using BOAST}

    \subsubsection{BigDFT}

BigDFT~\cite{bigdft} is a scientific application computing the electronic
structure of systems using density functional
theory (DFT). The application is mainly written in Fortran and currently
includes some 193,000 lines of Fortran code, which accounts for more than 50 \% 
% VANIA la plupart c'est Fortran et ensuite juste un peu plus de 50%, c'est quoi le reste?
of the code base. It is a parallel application based on the standards
MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also supports CUDA~\cite{cuda} and
OpenCL~\cite{opencl}.  As it is used on systems that may
have very different architectures, it is of great importance to be able to
optimize and run the application according to the specific underlying platform.

BigDFT is characterized by its extensive use of
convolution operators~\cite{nussbaumer1982fast} applied on large arrays of
data.  One example of a specific convolution, called
MagicFilter~\cite{Genovese|CAS2010}, can be seen in Listing~\ref{lst:conv_example}. 
It applies a filter $filt$ to the data set $in$ and then stores the result in the data set $out$ with a
transposition~\cite{Goedecker1993}.

\lstset{	language=C, 
		basicstyle=\scriptsize, 
		backgroundcolor=\color{white}, 
		frame=single, 
		captionpos=b,
		caption={MagicFilter}, 
		label={lst:conv_example}}

\begin{lstlisting}
double filt[16] = {F0, F1, ... , F15};
void magicfilter(int n, int ndat, double* in, double* out){
  double temp;
  int m;
  for( j = 0; j < ndat; j++) {
    for( i = 0; i < n; i++) {
      temp = 0;
      for( k = 0; k < 16; k++) {
        m = (i-7+k)%n
        temp += in[m + j*n] * filt[k];
      }
      out [j + i*ndat] = temp ;
    } 
  }
} 
\end{lstlisting}

As we can see, there are three nested loops working on arrays whose sizes
vary.  Various optimizations can be applied to this treatment and
may focus on the loop structure, as well as on the size of the data.  In this
paper we focus on this particular convolution and explore different
optimizations whose nature is presented in detail in the next section.

  \subsection{Porting SPECFEM3D to OpenCL using BOAST}

    \subsubsection{SPECFEM3D}
